\section{Appendix: Commands and Checklists}
\label{sec:AppendixCommands}

\subsection{Repository build commands}

\begin{longtable}{p{58mm}p{96mm}}
\toprule
Command & Purpose \\
\midrule
\shellcmd{mvn -q test} & Run Java tests in the Maven reactor. \\
\shellcmd{mvn -q -DskipJepsenTests=true test} & Run Java tests while skipping Jepsen integration. \\
\shellcmd{mvn -q -DskipTests package} & Build packages, source jars and Javadoc jars. \\
\shellcmd{cmake -S graft-cpp -B graft-cpp/build} & Configure the C++ build. \\
\shellcmd{cmake --build graft-cpp/build} & Build C++ targets. \\
\shellcmd{ctest --test-dir graft-cpp/build --output-on-failure} & Run C++ tests. \\
\shellcmd{cd graft-rust; cargo build -p graft-app-kv} & Build the Rust CLI binary. \\
\shellcmd{cd graft-rust; cargo test -p graft-tests} & Run Rust integration tests. \\
\shellcmd{cd graft-rust; cargo check} & Type-check the full Rust workspace. \\
\shellcmd{graft-cpp/run-mixed-suite.sh} & Run mixed Java/C++ smoke checks. \\
\shellcmd{graft-rust/tests/scripts/run-mixed-suite.sh} & Run mixed Java/Rust smoke checks. \\
\bottomrule
\end{longtable}

\subsection{Application developer checklist}

\begin{enumerate}
\item Define command and query payload formats.
\item Implement deterministic command application.
\item Implement application snapshot and restore.
\item Decide whether command results are needed.
\item Decide whether local queries are needed.
\item Add admission/authentication/authorization policy where appropriate.
\item Configure durable Raft storage.
\item Test leader, follower redirect, retry and invalid-payload behavior.
\item Test restart from log and snapshot.
\item Test membership changes if the application will operate dynamic clusters.
\end{enumerate}

\subsection{Raft machinery developer checklist}

\begin{enumerate}
\item Identify whether the change affects safety, liveness, protocol or packaging only.
\item Add focused Java and/or C++ unit tests for the affected rule.
\item Check Java/C++ status vocabulary and response shapes.
\item Check stable and joint-consensus majority behavior.
\item Check read-barrier behavior if reads are involved.
\item Check snapshot metadata and restore behavior if compaction or install is involved.
\item Run mixed smoke tests for cross-language protocol changes.
\item Run Jepsen when failures, partitions, membership or linearizable reads are involved.
\end{enumerate}

\subsection{Common failure interpretations}

\begin{description}
\item[Leader rejects write] The contacted node may be follower, learner,
decommissioned or unable to commit.
\item[Query returns retry] The leader could not refresh a quorum-backed read barrier.
\item[C++ Boost.Asio result\_of error] Older Boost.Asio was compiled as C++20
without the compatibility macro.
\item[Protobuf compiler/library warning] CMake found \texttt{protoc} and
\texttt{libprotobuf} with version strings that do not compare cleanly, often
because package manager prefixes differ.
\item[Mixed smoke status mismatch] Java and C++ disagree at the protocol
contract. Treat this as a parity defect unless explicitly domain-specific.
\end{description}
