\section{Jepsen Testing}
\label{sec:JepsenTesting}

Jepsen is the system-level fault-injection and linearizability verification
framework. It exercises the cluster under hostile conditions: process crashes,
network partitions, clock skew, membership changes, and snapshot boundary
exercises. This section describes the Jepsen architecture so machinery
developers can add new scenarios and debug failures.

\subsection{Architecture}

The Jepsen harness lives under \filepath{jepsen/}. It is a Clojure project
that orchestrates local OS processes or Docker containers, generates
randomized workloads, injects faults through a \emph{nemesis}, and checks
results against a Knossos linearizability model.

Key source files:

\begin{description}
\item[\filepath{src/raft\_jepsen/core.clj}] CLI parsing, test map construction,
generator, checker, and workload definitions.
\item[\filepath{src/raft\_jepsen/db.clj}] Node lifecycle: start, stop, kill, wipe,
port resolution, and implementation-specific command-lines (Java, C++, Rust).
\item[\filepath{src/raft\_jepsen/client.clj}] Workload executor: translates Jepsen
operations (\texttt{:write}, \texttt{:read}, \texttt{:cas}) into CLI commands,
parses responses, and classifies outcomes.
\item[\filepath{src/raft\_jepsen/nemesis.clj}] Fault injection: crash-restart,
process-pause, clock-skew, persistence-loss, network partitions, snapshot-boundary
exercises, and membership changes.
\item[\filepath{src/raft\_jepsen/observer.clj}] Observability: captures telemetry
and cluster-summary snapshots at interesting points for offline analysis.
\end{description}

\subsection{Node implementations}

The \texttt{--node-impls} flag controls which implementation each Jepsen node
uses. For a three-node mixed cluster:

\begin{lstlisting}[language=bash,caption={Mixed Jepsen run}]
cd jepsen
clojure -M:run mixed-all-three-java-leader \
  --time-limit 8 --concurrency 4 --node-count 3 \
  --node-impls java,cpp,rust \
  --client-impl mixed \
  --cpp-bin ../graft-cpp/build/graft_smoke \
  --rust-bin ../graft-rust/target/debug/graft-kv
\end{lstlisting}

Each node's implementation is determined by its position in the
\texttt{--node-impls} list. The \texttt{--client-impl mixed} flag means each
Jepsen worker uses the CLI matching its target node's implementation.

\subsection{Workload model}

Jepsen generates operations against a CAS (compare-and-set) register model.
The register can hold one value at a time; CAS compare-and-swap means
``if the current value is X, set it to Y.'' Knossos checks that every
read returns a value from a linearizable history.

Three operation types are generated:

\begin{description}
\item[\texttt{:write}] Put a key to a value (sent as \texttt{client-put}).
\item[\texttt{:read}] Read the current value of a key (sent as \texttt{client-get}).
\item[\texttt{:cas}] Conditionally write: supply expected and new values
(sent as \texttt{client-cas}).
\end{description}

\subsection{Nemesis catalog}

Each nemesis mode exercises a different failure domain:

\begin{longtable}{p{42mm}p{112mm}}
\toprule
Nemesis & Behavior \\
\midrule
\texttt{crash-restart} & Kill a node and restart it after a configurable interval. \\
\texttt{process-pause} & Send SIGSTOP/SIGCONT to freeze and resume a node process. \\
\texttt{clock-skew} & Inject a clock offset into a node's Raft time source. \\
\texttt{persistence-loss-restart} & Wipe a node's durable state and restart, checking catch-up. \\
\texttt{partition-one} & Isolate one node from the network. \\
\texttt{partition-leader} & Isolate the current leader. \\
\texttt{partition-leader-minority} & Isolate the leader plus one follower (minority partition). \\
\texttt{snapshot-boundary-restart} & Write a deterministic prelude, crash a node, advance the
log past the snapshot threshold, then restart the node. \\
\texttt{snapshot-creation-crash} & Cyclic crash of a follower during snapshot creation. \\
\texttt{snapshot-transfer-crash} & Kill/restart cycle during InstallSnapshot transfer. \\
\texttt{membership-join-promote} & Admit a new learner, promote to voter, finalize. \\
\texttt{membership-demote} & Demote a voter to learner. \\
\texttt{membership-remove-follower} & Remove a follower from the configuration. \\
\texttt{membership-remove-leader} & Remove the current leader. \\
\bottomrule
\end{longtable}

\subsection{Adding a nemesis}

To add a new nemesis:

\begin{enumerate}
\item Define a nemesis mode string in \filepath{core.clj} under the allowed
nemesis set and snapshot defaults.
\item Implement the nemesis logic in \filepath{nemesis.clj}:
  \begin{itemize}
  \item \texttt{invoke!} --- apply the fault to a node or the network,
  \item \texttt{teardown!} --- restore normal operation,
  \item \texttt{setup!} --- optional pre-fault preparation.
  \end{itemize}
\item Register the mode in \filepath{core.clj}'s \texttt{normalize-opts}
so snapshot-min-entries and snapshot-chunk-bytes are set if needed.
\item Add test cases to \filepath{run-suite.sh} with appropriate port bases.
\item Run the new nemesis in isolation first, then as part of the extended
or stress suite.
\end{enumerate}

\subsection{Reading failure output}

Jepsen produces several artifacts in the work directory:

\begin{description}
\item[\filepath{history.edn}] The full operation history with timing and outcomes.
\item[\filepath{results.edn}] The checker verdict (\texttt{:valid? true/false})
and per-worker statistics.
\item[\filepath{timeline.html}] A visual timeline of operations and nemesis actions.
\item[\filepath{store/}] Jepsen's own durable store for the checker.
\item[\filepath{observations/}] Observer-captured JSONL files with telemetry
and cluster-summary snapshots at failure points.
\end{description}

When a test fails, check the \texttt{client-anomaly} and
\texttt{telemetry-parse-failure} entries in the observer JSONL for the
cluster state at the moment of failure. The Jepsen log
(\filepath{jepsen.log}) contains the full process output.

\subsection{Suite structure}

The test suites in \filepath{run-suite.sh} are organized by confidence level:

\begin{description}
\item[\texttt{smoke}] Fast baseline tests with no nemesis (port bases 10080--10180).
\item[\texttt{extended}] 14 cases covering all nemesis modes (port bases 10080--21420).
\item[\texttt{mixed}] 25 cases covering Java and C++, Java and Rust, and three-way
Java and C++ and Rust clusters (port bases 21480--24130).
\item[\texttt{stress}] 9 cases with 300-second runs, concurrency 20, 5000 operations
(port bases 22580--23380).
\end{description}

\subsection{Docker/SRV Jepsen}

The Jepsen harness also supports a Docker backend (\texttt{--backend docker-srv})
that orchestrates clusters through Docker Compose instead of spawning local
OS processes. This is useful for testing DNS SRV-based bootstrapping and for
reproducing failures in a containerized environment.

Supported nemeses in Docker mode are limited to \texttt{crash-restart},
\texttt{process-pause}, and \texttt{partition-*} (network disconnect/connect
via \texttt{docker network}). Filesystem-based nemeses (persistence-loss,
snapshot-boundary) and clock-skew are not supported in Docker mode because
they require direct host filesystem access or process-level clock adjustment.
