\section{Rust Library}
\label{sec:RustLibrary}

The Rust implementation lives under \filepath{graft-rust}. It is a Cargo
workspace built with \texttt{cargo} so Rust deployments do not depend on the
Maven reactor or CMake. It shares the protobuf schema and envelope protocol with
Java and C++.

\subsection{Build model}

The Rust workspace contains seven crates:

\begin{itemize}
\item \texttt{graft-proto}: shared protobuf generated types plus varint32
envelope framing (byte-compatible with the Java
\texttt{ProtobufLiteEncoder}/\texttt{ProtobufLiteDecoder} and the C++
\texttt{envelope\_codec.hpp}),
\item \texttt{graft-core}: transport-agnostic Raft consensus engine (elections,
log replication, commit/apply, snapshots, joint-consensus membership, read
barriers),
\item \texttt{graft-storage}: in-memory and file-based log and persistent state
stores,
\item \texttt{graft-transport}: async TCP client and server with varint-length-prefixed
envelope framing,
\item \texttt{graft-runtime}: active event loop (\texttt{RaftRuntime}), message
handler, membership encoders, seed providers, policy hooks,
\item \texttt{graft-app-kv}: demo key-value state machine and the
\texttt{graft-kv} CLI binary,
\item \texttt{graft-tests}: integration tests covering elections, log
replication, snapshots, membership, and file persistence.
\end{itemize}

Basic workflow:

\begin{lstlisting}[language=bash,caption={Rust build workflow}]
cd graft-rust
cargo build -p graft-app-kv     # build the CLI binary
cargo test -p graft-tests       # run the integration tests
cargo check                     # type-check the entire workspace
\end{lstlisting}

The code targets stable Rust (edition 2021). The build fetches crate dependencies
automatically from crates.io. The shared \filepath{raft-wire/src/main/proto/raft.proto}
is compiled by the \texttt{graft-proto} crate's \texttt{build.rs} using
\texttt{prost-build}.

\subsection{Application boundary}

The Rust application boundary mirrors the Java state-machine boundary:

\begin{itemize}
\item \texttt{StateMachine} trait with \texttt{apply}, \texttt{snapshot}, and
\texttt{restore},
\item \texttt{QueryableStateMachine} extension for read queries,
\item \texttt{ResultStateMachine} extension for client-visible results,
\item \texttt{LogStore} and \texttt{PersistentStateStore} traits for storage
backends.
\end{itemize}

The demo \texttt{KeyValueStateMachine} implements \texttt{StateMachine} and
\texttt{QueryableStateMachine}. Commands and queries are JSON-serialized into the
Raft log entry data field. Domain application developers should implement these
traits and register their state machine with a \texttt{RaftNode}.

\subsection{Runtime and transport}

Tokio provides async client and server transport over the shared envelope
protocol. The \texttt{RaftRuntime} is the active event loop: it runs
randomized election timers (1500--3000\,ms), broadcasts heartbeats every 750\,ms,
and processes responses to update \texttt{matchIndex}/\texttt{nextIndex} and
advance the commit index.

The \texttt{RaftHandler} dispatches all 11 Raft RPC types and enforces
configurable authentication, authorization, admission, and rate-limiting policies
through the policy traits in \texttt{graft-runtime/src/policy.rs}.

The active server modes wire a local Rust \texttt{RaftNode} to the RPC handler
and transport server. The smoke CLI can also act as a client for Java, C++, or
Rust peers.

Use Rust smoke commands to validate boundary behavior:

\begin{lstlisting}[language=bash,caption={Selected Rust probe commands}]
graft-rust/target/debug/graft-kv client-put 127.0.0.1 11082 k v1 rust-client
graft-rust/target/debug/graft-kv client-get 127.0.0.1 11082 k rust-client
graft-rust/target/debug/graft-kv cluster-summary 127.0.0.1 11082
graft-rust/target/debug/graft-kv reconfiguration-status 127.0.0.1 11082
\end{lstlisting}

\subsection{Cross-language mixed smoke tests}

Rust↔Java mixed smoke scripts mirror the existing C++↔Java scripts:

\begin{lstlisting}[language=bash,caption={Rust mixed smoke tests}]
graft-rust/tests/scripts/run-mixed-smoke-rust-leader.sh
graft-rust/tests/scripts/run-mixed-smoke-java-leader.sh
graft-rust/tests/scripts/run-interop-smoke.sh
graft-rust/tests/scripts/run-mixed-suite.sh
\end{lstlisting}

These reuse the Java probe from \filepath{graft-cpp/java-probe} and exercise
elections, log replication, telemetry, and client redirects across Rust and Java
nodes. Run these after changing protobuf messages, status strings, or any RPC
behavior.

\subsection{Protobuf toolchain}

The Rust code uses the same shared \texttt{raft.proto} as Java and C++. The
\texttt{prost} crate generates Rust types at build time. No manual code
generation step is needed---it runs as part of \texttt{cargo build}. The
generated types use Rust protobuf encoding that is byte-compatible with Java
protobuf-lite and C++ libprotobuf.

\subsection{Policy hooks}

The Rust runtime includes pluggable policy hooks matching the Java
\texttt{raft-runtime} interfaces:

\begin{itemize}
\item \texttt{Authenticator} --- validates client credentials (default:
\texttt{AllowAllAuthenticator}, also includes \texttt{SharedSecretAuthenticator}),
\item \texttt{Authorizer} --- authorizes commands against a principal (default:
\texttt{AllowAllAuthorizer}),
\item \texttt{AdmissionPolicy} --- decides accept/redirect/reject for writes
(default: \texttt{AcceptAllAdmissionPolicy}, also includes
\texttt{LeaderRedirectAdmissionPolicy}),
\item \texttt{RateLimiter} --- sliding-window rate limiter for telemetry and
cluster-summary requests,
\item \texttt{StuckDetector} --- reports prolonged joint-consensus transitions.
\end{itemize}

Application modules can replace the default policy components by setting the
\texttt{Arc} fields on \texttt{RaftHandler}.

\subsection{Docker support}

A multi-stage Dockerfile builds the Rust binary inside a \texttt{rust:1-bookworm}
container and copies it into a slim \texttt{debian:bookworm-slim} runtime image.
Compose files for Rust-only and mixed Java+C+++Rust clusters are available:

\begin{lstlisting}[language=bash,caption={Docker Rust modes}]
docker/run-srv-smoke.sh rust          # 3 Rust nodes
docker/run-srv-smoke.sh mixed         # 1 Java + 1 C++ + 1 Rust
docker/run-srv-suite.sh               # all 4 modes (java, cpp, rust, mixed)
\end{lstlisting}

Nodes bootstrap from explicit peer lists (\texttt{RAFT\_PEERS}) or DNS SRV
records (\texttt{RAFT\_CLUSTER\_SRV}) via a shell entrypoint.

\subsection{Parity expectations}

Rust should be judged against Java and C++ at the Raft behavior boundary:

\begin{itemize}
\item status values such as \texttt{ACCEPTED}, \texttt{REDIRECT},
\texttt{RETRY} and \texttt{NOT\_LEADER},
\item vote and append semantics,
\item snapshot install and restore semantics,
\item read barrier and read-lease behavior,
\item joint-consensus majority rules,
\item membership admission and finalization,
\item mixed Java/C++/Rust cluster behavior.
\end{itemize}

Differences in domain examples can be tolerated temporarily. Differences in Raft
safety semantics should be treated as defects.
