\section{System Overview}
\label{sec:SystemOverview}

The repository contains one shared protocol and three implementation tracks:

\begin{description}
\item[Java] Maven modules implementing the main Raft runtime, storage, transport,
state-machine contracts, example applications, telemetry and tests.
\item[C++] A CMake subtree under \filepath{graft-cpp} implementing the same wire
protocol and a bounded but growing Raft runtime using Boost.Asio.
\item[Rust] A Cargo workspace under \filepath{graft-rust} implementing the same
wire protocol and a functionally equivalent Raft runtime using Tokio.
\item[Shared wire schema] Protobuf messages under \filepath{raft-wire/src/main/proto}
used by Java, C++ and Rust.
\item[Jepsen] Workload and fault-injection tests under \filepath{jepsen}, including
mixed Java/C++/Rust scenarios.
\end{description}

\subsection{Java module map}

\begin{longtable}{p{42mm}p{112mm}}
\toprule
Module & Responsibility \\
\midrule
\filepath{raft-wire} & Protocol model, protobuf mapping and shared wire messages. \\
\filepath{raft-core} & Raft node state machine: terms, votes, log matching, commit,
snapshot handling, read safety and membership application. \\
\filepath{raft-membership} & Cluster configuration model and joint-consensus
majority calculations. \\
\filepath{raft-state-machine} & Application state-machine interfaces:
\texttt{SnapshotStateMachine}, \texttt{QueryableStateMachine} and result-returning variants. \\
\filepath{raft-storage} & In-memory and file-backed persistent state and log stores. \\
\filepath{raft-transport-netty} & Netty transport client/server for Raft and client RPCs. \\
\filepath{raft-runtime} & Application bootstrap, adapter, operational handlers and
policy hooks. \\
\filepath{raft-app-kv} & Basic key-value demo application. \\
\filepath{raft-app-reference} & Reference-data application example and policy behavior. \\
\filepath{raft-telemetry} & Telemetry exporter support. \\
\filepath{raft-dist} & Runnable distribution entry point. \\
\filepath{raft-tests} & Unit, integration and cross-module tests. \\
\bottomrule
\end{longtable}

\subsection{C++ layout}

The C++ subtree is intentionally smaller but follows the same conceptual layers:

\begin{description}
\item[\filepath{include/graft/core}] Raft node, application state-machine and snapshot codec.
\item[\filepath{include/graft/runtime}] Runtime and RPC handler abstractions.
\item[\filepath{include/graft/storage}] Persistent state store.
\item[\filepath{include/graft/transport}] Boost.Asio client/server transport.
\item[\filepath{include/graft/telemetry}] Prometheus, OTLP and OTEL JSON log export.
\item[\filepath{src}] Implementations of the core, runtime, storage, transport, telemetry and CLI.
\item[\filepath{tests}] Catch2 tests for the C++ core and runtime behaviors.
\end{description}

\subsection{Rust layout}

The Rust workspace is structured as seven crates mirroring the Java module boundaries:

\begin{description}
\item[\filepath{graft-proto}] Generated protobuf types and varint32 envelope codec.
\item[\filepath{graft-core}] Transport-agnostic Raft consensus engine.
\item[\filepath{graft-storage}] In-memory and file-based log and persistent state stores.
\item[\filepath{graft-transport}] Tokio-based TCP client/server with envelope framing.
\item[\filepath{graft-runtime}] Active event loop, RPC handler, membership encoders, seed providers, policy hooks.
\item[\filepath{graft-telemetry}] Prometheus scrape endpoint, OTLP push and OTEL JSON log export.
\item[\filepath{graft-app-kv}] Demo key-value state machine and the \texttt{graft-kv} CLI binary.
\item[\filepath{graft-tests}] Integration tests covering elections, replication, snapshots, membership, and persistence.
\end{description}

\subsection{Boundary principle}

The central design rule is:

\begin{quote}
Raft owns replication safety. The application owns command meaning.
\end{quote}

That has practical consequences:

\begin{itemize}
\item Raft persists terms, votes, log entries, commit index, membership and snapshot metadata.
\item The application persists its own state only through deterministic command application and application snapshot bytes.
\item Cluster configuration is part of Raft state, not part of the domain model, although it is persisted beside the replicated application data path.
\item Application code should not decide majorities, refresh read leases, send vote requests or mutate the Raft log directly.
\end{itemize}

\subsection{Shared protocol}

Java, C++ and Rust interoperate because they share:

\begin{itemize}
\item protobuf message definitions,
\item envelope framing,
\item client command and query request/response shapes,
\item membership and join/reconfiguration messages,
\item telemetry and cluster-summary messages,
\item snapshot install messages.
\end{itemize}

Cross-language compatibility should be tested at the protocol boundary. If Java,
C++ or Rust disagree about status strings, read-safety conditions, membership
majorities or snapshot metadata, that is a library bug unless the difference is
explicitly documented as domain-specific.
