\section{Testing and Operations}
\label{sec:TestingAndOperations}

\subsection{Test layers}

The project uses several layers of verification:

\begin{description}
\item[Java unit tests] Core election, append, snapshot, membership, read-lease,
adapter and storage behavior.
\item[Java transport tests] Netty client/server behavior and response handling.
\item[C++ unit tests] C++ core, membership, state-machine, snapshot and RPC behavior.
\item[Rust integration tests] Rust core, membership, state-machine, snapshot,
replication and file persistence behavior.
\item[Mixed smoke tests] Java/C++/Rust interoperability at the shared wire boundary.
\item[Jepsen] Fault-injection and workload checks for stronger system-level confidence.
\end{description}

Use the smallest test layer that can prove a change, then add broader tests when
the change crosses module, language or safety boundaries.

\subsection{Java test workflow}

\begin{lstlisting}[language=bash,caption={Java test workflow}]
mvn -q test
mvn -q -DskipJepsenTests=true test
mvn -q -DskipTests package
\end{lstlisting}

Unit tests should avoid real Netty resources unless the test is explicitly about
transport behavior. Use no-op or in-memory transport fakes for Raft core tests so
older platforms with lower file-descriptor limits do not fail while constructing
test nodes.

\subsection{C++ test workflow}

\begin{lstlisting}[language=bash,caption={C++ test workflow}]
cmake -S graft-cpp -B graft-cpp/build
cmake --build graft-cpp/build
ctest --test-dir graft-cpp/build --output-on-failure
\end{lstlisting}

When CMake finds the wrong protobuf tools, pin \texttt{protoc}. If Boost.Asio
fails around removed standard-library traits, check the C++ compatibility macro
described in Section~\ref{sec:CppLibrary}.

\subsection{Rust test workflow}

\begin{lstlisting}[language=bash,caption={Rust test workflow}]
cd graft-rust
cargo test -p graft-tests          # integration tests
cargo test --workspace             # all tests across all crates
cargo build -p graft-app-kv        # build the CLI binary
\end{lstlisting}

The integration tests use in-memory stores and exercise elections, log
replication, snapshot install, joint-consensus membership transitions,
and file-based persistence round-tripping. No external dependencies are required
for tests---the entire suite runs offline.

\subsection{Mixed Java/C++/Rust checks}

The mixed smoke scripts validate that all implementations agree on the shared
protocol:

\begin{lstlisting}[language=bash,caption={Mixed implementation checks}]
graft-cpp/run-mixed-smoke.sh                         # C++ leader + Java follower
graft-cpp/run-mixed-smoke-java-leader.sh             # Java leader + C++ follower
graft-cpp/run-mixed-membership-smoke.sh              # membership with C++ learner
graft-cpp/run-mixed-membership-cpp-leader-smoke.sh   # membership with C++ leader
graft-cpp/run-mixed-snapshot-smoke.sh                # snapshot catch-up
graft-cpp/run-mixed-suite.sh                         # all C++ mixed scenarios
graft-rust/tests/scripts/run-mixed-suite.sh          # Rust↔Java smoke suite
\end{lstlisting}

Run these after changing:

\begin{itemize}
\item protobuf messages,
\item status strings,
\item client command/query behavior,
\item membership behavior,
\item read-barrier behavior,
\item snapshot install or snapshot metadata.
\end{itemize}

\subsection{Jepsen}

Jepsen should be used when a change affects safety under concurrency, failure or
network partitions. In particular, run or update Jepsen scenarios for:

\begin{itemize}
\item leader election changes,
\item commit-index changes,
\item read-lease/read-barrier changes,
\item membership changes,
\item mixed Java/C++/Rust cluster behavior,
\item client workload semantics such as CAS.
\end{itemize}

The Jepsen suite is not a replacement for unit tests. It is the system-level
check that the library still behaves correctly when timing and partitions are
hostile.

\subsection{Operational notes}

A Raft deployment needs durable per-node storage. Losing local Raft storage can
lose votes, terms, log entries and membership state. For application developers,
this means the storage path is part of the cluster deployment contract, even if
the domain state itself looks like an in-memory hashtable.

Operators should monitor:

\begin{itemize}
\item current leader,
\item term,
\item commit index and last applied index,
\item follower replication lag,
\item quorum health,
\item current and next quorum health during joint consensus,
\item stuck reconfiguration age,
\item snapshot install events.
\end{itemize}

\subsection{Cross-language parity}

The three implementations (Java, C++, Rust) are intended to be functionally
equivalent at the Raft boundary. When changing one, verify the others against
this checklist:

\begin{itemize}
\item Status strings must be identical across implementations: \texttt{ACCEPTED},
\texttt{REDIRECT}, \texttt{RETRY}, \texttt{NOT\_LEADER}, \texttt{INVALID},
\texttt{UNAUTHORIZED}, \texttt{RATE\_LIMITED}.
\item Vote-granting rules (term comparison, log up-to-date check) must be
identical.
\item AppendEntries conflict resolution (backtracking, truncation, match-index
advancement) must be identical.
\item Snapshot install and restore semantics (chunk assembly, metadata wrapping)
must be identical.
\item Joint-consensus majority rules must use the same quorum formula in all
three.
\item Membership admission and finalization (JOIN↦JOINT↦FINALIZE) must be
identical.
\item Read barriers must contact a joint-consensus quorum of voters before
serving.
\item Client redirect behavior (leader-id, leader-host, leader-port in
responses) must be identical.
\item Differences in domain examples (key-value store vs. reference-data) are
acceptable.
\end{itemize}

\subsection{Codebase navigation}

For machinery developers who need to find specific Raft behaviors across
implementations:

\begin{longtable}{p{48mm}p{36mm}p{36mm}p{36mm}}
\toprule
Behavior & Java & C++ & Rust \\
\midrule
Election timeout / jitter & \filepath{raft-core/.../RaftNode.java} &
\filepath{graft-cpp/src/core/raft\_node.cpp} &
\filepath{graft-rust/graft-core/src/raft\_node.rs} \\
\midrule
AppendEntries follower & RaftNode.java &
raft\_node.cpp & raft\_node.rs \\
\midrule
Commit advancement & RaftNode.java &
raft\_node.cpp & raft\_node.rs \\
\midrule
Snapshot install (follower) & RaftNode.java &
raft\_node.cpp & raft\_node.rs \\
\midrule
Joint-consensus majority & \filepath{raft-membership/.../ClusterConfiguration.java} &
\filepath{graft-cpp/src/core/cluster\_membership.cpp} &
\filepath{graft-rust/graft-core/src/membership.rs} \\
\midrule
Active runtime loop & \filepath{raft-runtime/.../BasicAdapter.java} &
\filepath{graft-cpp/src/runtime/raft\_runtime.cpp} &
\filepath{graft-rust/graft-runtime/src/runtime.rs} \\
\midrule
RPC handler dispatch & \filepath{raft-transport-netty/.../RaftMessageHandler.java} &
\filepath{graft-cpp/src/runtime/rpc\_handler.cpp} &
\filepath{graft-rust/graft-runtime/src/handlers.rs} \\
\midrule
Transport framing & \filepath{raft-transport-netty/.../ProtobufLiteEncoder.java} &
\filepath{graft-cpp/include/graft/wire/envelope\_codec.hpp} &
\filepath{graft-rust/graft-proto/src/lib.rs} \\
\midrule
Membership encoders & ClusterConfigurationCommand.java &
rpc\_handler.cpp (inline) &
\filepath{graft-rust/graft-runtime/src/membership.rs} \\
\midrule
Auth / admission hooks & \filepath{raft-runtime/.../ClientWriteAdmissionPolicy.java} &
rpc\_handler.cpp (Authenticator) &
\filepath{graft-rust/graft-runtime/src/policy.rs} \\
\midrule
State machine interface & \filepath{raft-state-machine/.../SnapshotStateMachine.java} &
\filepath{graft-cpp/include/graft/core/application\_state\_machine.hpp} &
\filepath{graft-rust/graft-core/src/state\_machine.rs} \\
\midrule
Example KV application & \filepath{raft-app-kv/} &
\filepath{graft-cpp/src/core/key\_value\_state\_machine.cpp} &
\filepath{graft-rust/graft-app-kv/src/lib.rs} \\
\bottomrule
\end{longtable}

Changes that affect Raft behavior should be made in all three files at the same
row of this table. Unit tests and mixed-smoke scripts should be run for each
affected implementation.
