\section{Wire, Storage and Snapshots}
\label{sec:WireStorageSnapshots}

\subsection{Wire protocol}

All cross-node and client-facing transport uses a shared protobuf schema. The
transport envelope carries:

\begin{itemize}
\item correlation id,
\item term,
\item message type,
\item protobuf payload.
\end{itemize}

Java maps protobuf messages through the \texttt{ProtoMapper}. C++ generates
protobuf classes from the same schema and uses the same envelope framing. When a
new message is added, both languages must be updated and mixed smoke coverage
should be considered.

\subsection{Status vocabulary}

Status strings are part of the cross-language contract. Examples include:

\begin{description}
\item[\texttt{ACCEPTED}] A client command was accepted after commit/apply.
\item[\texttt{REDIRECT}] The contacted node is not the leader and can identify a leader.
\item[\texttt{RETRY}] The request may be retried, commonly after read-barrier failure.
\item[\texttt{INVALID}] The request payload or operation is invalid.
\item[\texttt{UNAUTHORIZED}] Authentication or authorization failed.
\end{description}

Avoid introducing language-specific synonyms. If Java says \texttt{ACCEPTED},
C++ should not say \texttt{OK} for the same behavior.

\subsection{Persistent state}

Raft persistence includes:

\begin{itemize}
\item current term,
\item voted-for candidate,
\item known leader where applicable,
\item log entries,
\item snapshot boundary and metadata,
\item cluster configuration.
\end{itemize}

Application state is persisted indirectly by applying committed entries and by
snapshotting application state. If a node restarts, it must reconstruct both Raft
metadata and application state consistently.

\subsection{Snapshot wrapping}

Snapshots have two layers:

\begin{description}
\item[Application payload] Bytes produced by the state machine.
\item[Raft wrapper] Metadata such as last included index, last included term and
cluster configuration at that boundary.
\end{description}

This wrapper is essential for restart and catch-up. A follower receiving a
snapshot must restore application state and update Raft membership/log boundary
state consistently.

\subsection{Configuration storage}

Cluster configuration is not an external domain object. It is part of the Raft
log and snapshot metadata. The application may expose cluster configuration
through operational APIs, but it should not duplicate configuration as domain
state unless it is doing so for audit or display.

\subsection{Protocol evolution}

When changing the shared protocol:

\begin{enumerate}
\item Update the protobuf schema.
\item Regenerate Java and C++ bindings.
\item Update Java mapping and C++ mapping/handling.
\item Add unit tests in both languages where behavior changes.
\item Add mixed smoke or Jepsen coverage when the change affects cross-language behavior.
\end{enumerate}

Prefer additive schema changes. If a field meaning changes, document the version
expectations and test mixed-version behavior before relying on it operationally.
